Performance Models for Data Transfers: A Case Study with Molecular Chemistry Kernels

### 2019-04-05 ICPP-19/communications.tex
ICPP-19 format

%
\documentclass[runningheads]{llncs} %
\usepackage{graphicx}

@@

\usepackage{algorithmic}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage[cmex10,fleqn]{amsmath}
\usepackage{amsfonts}

@@

\newcommand{\scomm}{\ensuremath{{\cal S}_{\text{COMM}}}}
\newcommand{\scomp}{\ensuremath{{\cal S}_{\text{COMP}}}}

@@

\begin{document} %
	%	\title{Contribution Title\thanks{Supported by organization x.}}
	\title{Performance Models for Data Transfers: A Case Study with Molecular Chemistry Kernels}
	%\titlerunning{Abbreviated paper title}
	% If the paper title is too long for the running head, you can set
	% an abbreviated paper title here
	%
	\author{}
%%%%	\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and Second
%%%%		Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and Third
%%%%		Author\inst{3}\orcidID{2222--3333-4444-5555}} %
%%%%	\authorrunning{F. Author et al.} % First names are abbreviated in the running
	% head.
	% If there are more than two authors, 'et al.' is used.
	%
	\institute{Princeton University, Princeton NJ 08544, USA \and Springer
		Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
		\email{lncs@springer.com}\\
		\url{http://www.springer.com/gp/computer-science/lncs} \and ABC Institute,
		Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
		\email{\{abc,lncs\}@uni-heidelberg.de}} %
	\maketitle              % typeset the header of the contribution
	%
	\begin{abstract} 
%%		
%%		The abstract should briefly summarize the contents of the
%%		paper in 150--250 words.

With increasing complexity of hardwares, systems with different memory nodes are ubiquitous in high performance computing (HPC). It is paramount to develop strategies to overlap the data transfers between memory nodes with the computation in order to exploit the full potential of these systems. In this article, we consider the problem of deciding the  order of data transfers between two memory nodes for a set of independent tasks  with the objective to minimize the makespan. We propose an optimal heuristic to obtain the order of data transfers when there is not any restriction on memory limit. We also show that with finite memory capacity obtaining the optimal order of data transfers is an NP-complete problem. We propose several heuristics for this problem and provide details about their favorable situations. We present analysis of our heuristics on traces, obtained by running 2 molecular chemistry kernels, namely, Hartree–Fock (HF)  and Coupled Cluster Single Double (CCSD) on 10 nodes of an HPC system. Our results show that some of our heuristics achieve significant overlap for moderate memory capacities and are very close to the lower bound of makespan.

		\keywords{Communication Scheduling  \and Memory Nodes \and Runtime Systems \and Communication-Computation Overlap \and Molecular Chemistry.}
	\end{abstract} %

@@

	\begin{figure}[htb]
		\centering
		\includegraphics[width=\textwidth, height=0.5\textheight, 
		keepaspectratio]{Figs/swappedSchedule.jpg}
		\caption{ \label{fig:bothSchedule} Original and Swapped Schedule}
	\end{figure}

	We write the following constraints based on the two schedules of the Fig~\ref{fig:bothSchedule}.

@@

   		\begin{table*}[htb]
   			\begin{tabular}{ |c|c|c| }
   				\hline
   				Task & Input Transfer time ($comm_i$)& Computation time ($e_i$)\\ \hline
   				$K_0$ & $0$ & $3$ \\ \hline
   				$K_1, \cdots, K_{m-1}$ & $b'=b+3x$ & $3$\\ \hline
   				$K_m$ & $b'=b+3x$ & $0$ \\ \hline
   				$1\le i \le 3m, A_i$ & $1$ & $a_i' = a_i +x$\\ \hline
   			\end{tabular}
   		\end{table*}
   	Where $x= 2 max\{a_i:1\le i\le 3m\}$\\ 
   	Memory capacity $C=b'+3$ and $L=m(b'+3)$, this completes the construction of instance $I$ of the 
   	problem $DT$.

@@

   	Conversely, if such a partition of $A$ is achievable, then we can construct a feasible schedule $S$ 
   	without idle times by the pattern depicted in Fig~\ref{fig:firstSegment}. We have shown that $3PAR$ 
   	problem reduces to $DT$, and hence problem $DT$ is NP-complete.

   	\begin{figure}[htb]
   		\centering
   		\includegraphics[scale=0.07]{Figs/3PARToSched.jpg}
   		\caption{ \label{fig:firstSegment} First segment of Schedule $S$}
   	\end{figure}

@@

\begin{table}[htb]
	\scriptsize
	\begin{tabular}{|c|p{3cm}|}
		\hline
		\textbf{Heuristic} & \textbf{Favorable Situation} \\ \hline
		order\_of\_optimal\_strategy\_infinite\_memory & Unlimited memory capacity (\textcolor{green}{Optimal}) \\ \hline
		increasing\_order\_of\_communication\_strategy & Unlimited memory capacity and tasks are compute intensive (\textcolor{green}{Optimal}) \\ \hline
		decreasing\_order\_of\_computation\_strategy & Unlimited memory capacity and tasks are communication intensive (\textcolor{green}{Optimal}) \\ \hline
		increasing\_order\_of\_communication\_plus\_computation\_strategy & Moderate memory capacity and most tasks are compute intensive \\ \hline
		decreasing\_order\_of\_communication\_plus\_computation\_strategy & Moderate memory capacity and most tasks are communication intensive \\ \hline
		largest\_communication\_task\_respects\_memory\_restriction &  Limited memory capacity and significant percentage of tasks are communication intensive \\ \hline
		smallest\_communication\_task\_respects\_memory\_restriction & Limited memory capacity and most tasks are compute intensive \\ \hline
		maximum\_accelerated\_task\_respects\_memory\_restriction &  Limited memory capacity and number of communication and compute intensive tasks are approximately equal \\ \hline
		optimal\_order\_infinite\_memory\_largest\_communication\_task\_respects\_memory\_restriction & Moderate memory capacity and significant percentage of tasks are communication intensive\\ \hline
		optimal\_order\_infinite\_memory\_smallest\_communication\_task\_respects\_memory\_restriction & Moderate memory capacity and most tasks are compute intensive \\ \hline
		optimal\_order\_infinite\_memory\_maximum\_accelerated\_task\_respects\_memory\_restriction & Moderate   memory capacity and number of communication and compute intensive tasks are approximately equal \\ \hline
	\end{tabular}\caption{~\label{tab:heuristicsAndFavorableScenarios}Heuristics and their favorable scenarios}
\end{table}

@@

	Figures ~\ref{fig:ratio_to_optimal_hf} and ~\ref{fig:ratio_to_optimal_ccsd} depict the distribution of the performance  of each heuristic for all considered memory capacities, where plots are categorized by memory capacities. For each memory capacity and each heuristic, the box on the plot displays the median, first and last quartile, and the whiskers indicate minimum and maximum values, with outliers are shown by black dots.
\subsection{HF Performance}	
	\begin{figure}[htb]
		\includegraphics[scale=0.225]{./results/plots/ratio_to_optimal_selected_hf.pdf}
		\caption{Comparison of different heuristics for HF.}
		\label{fig:ratio_to_optimal_hf}
	\end{figure}
 As indicated above, HF tasks are highly homogeneous, this is also noticeable in Figure~\ref{fig:ratio_to_optimal_hf}. All heuristics depict similar behavior for minimum memory capacity $m_c$  and  increasing the memory capacity slightly does not change the performance of all heuristics. As memory capacity increased further, dynamic variants of heuristics starts performing better and for the moderate memory capacities (close to $2m_c$ ), static order with dynamic correction variants outperform others. \textit{Gilmore-Gomory} heuristic does not achieve good performance. it can be understand from the fact that the task sequence is obtained considering no extra memory is available, but here we apply the same sequence for a different scenario,  overall memory consumed at each point is limited to memory capacity.

	\begin{figure}[htb]
	\includegraphics[scale=0.225]{./results/plots/inverse_ratio_to_optimal_hf-best.pdf}
	\caption{Comparision of best variants of all categories for HF.}
	\label{fig:ratio_to_optimal_best_hf}
	\end{figure}

@@

	\begin{figure}[htb]
	\includegraphics[scale=0.225]{./results/plots/ratio_to_optimal_selected_ccsd.pdf}
	\caption{Comparison of different heuristics for CCSD.}
	\label{fig:ratio_to_optimal_ccsd}
	\end{figure}

@@

	\begin{figure}[htb]
	\includegraphics[scale=0.225]{./results/plots/inverse_ratio_to_optimal_ccsd-best.pdf}
	\caption{Comparision of best variants of all categories for CCSD.}
	\label{fig:ratio_to_optimal_best_ccsd}
	\end{figure}

@@

\subsection{MILP Based Strategy}
 We use GLPK solver v4.65 to solve the MILP problem described in Section~\ref{subsec:linearprogrammingformulation}. The solver was unable to handle number of variables required to solve MILP at the scale of our interest. Hence, we solve the linear program iteratively for a small subset of size $k=3,4,5$ (solver was unable to handle number of variables required for $k\ge6$ ) and represent the makespan calculated by this heuristic as $lp.k$. At the boundary of two iterations we fixed the event (communication or computation) of an unfinished task who has  started before the boundary point and consider other events flexible. We compute various  $lp.k$ values for different memory capacities and observe that most of the other heuristics perform better than this heuristic. Figure~\ref{fig:iterativeLpSolution} shows the comparison of different heuristics with MILP based heuristic for various memory capacities of a single trace file.

	\begin{figure}[htb]
	\includegraphics[scale=0.5]{./results/makespan_with_lp.pdf}
	\caption{Comparision of different heuristics with MILP solution based heuristic.}
	\label{fig:iterativeLpSolution}
\end{figure}

@@

 \begin{figure}[htb]
	\subfloat[Best variants of HF.\label{fig:hf_best_variants_batch}]{%
		\includegraphics[width=.45\linewidth]{./results/plots/inverse_ratio_to_optimal_hf_batch-best.pdf}
	}%\\
	\subfloat[Best variants of CCSD.\label{fig:ccsd_best_variants_batch}]{%
		\includegraphics[width=.45\linewidth]{./results/plots/inverse_ratio_to_optimal_ccsd_batch-best.pdf}
	}%\\
	\caption{Best variants of all categories where heuristics are applied in the batches of 100 tasks.}
	\label{fig:best_variants_batch}
 \end{figure}

@@

\subsection{Impact of heterogeneity on different heuristics}
\todo[inline]{SKumar: remove this subsection or replace with an experiment which shows the performance difference with all classes of heuristics for randomly generated inputs.}
	\begin{figure}[htb]
	\includegraphics[scale=0.225]{./results/plots/ratio_to_optimal_random_selected_ccsd.pdf}
	\caption{Comparison of different heuristics for CCSD with highly heterogeneous tasks.}
	\label{fig:ratio_to_optimal_ccsd_random}
	\end{figure}

@@

\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{communications}

>>>>

\documentclass[sigconf]{acmart}

%%%
%%% \BibTeX command to typeset BibTeX logo in the docs
%%\AtBeginDocument{%
%%  \providecommand\BibTeX{{%
%%    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

\copyrightyear{2019}
\acmYear{2019}
\setcopyright{acmlicensed}
\acmConference[ICPP '19]{ICPP 2019: 48th International Conference on Parallel Processing}{August 05--08, 2019}{Kyoto, Japan}
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
%%\acmPrice{15.00}
%%\acmDOI{10.1145/1122445.1122456}
%%\acmISBN{978-1-4503-9999-9/18/06}

\usepackage{graphicx}

@@

\usepackage{algorithmic}
\usepackage[ruled,vlined]{algorithm2e}
%%\usepackage[cmex10,fleqn]{amsmath}
\usepackage{amsfonts}

@@

%%\newcommand{\scomm}{\ensuremath{{\cal S}_{\text{COMM}}}}
%%\newcommand{\scomp}{\ensuremath{{\cal S}_{\text{COMP}}}}

\newcommand{\scomm}{\ensuremath{{S}_{\text{COMM}}}}
\newcommand{\scomp}{\ensuremath{{S}_{\text{COMP}}}}

@@

\begin{document}

\title{Performance Models for Data Transfers: A Case Study with Molecular Chemistry Kernels}

\author{Valerie B\'eranger}
\affiliation{%
  \institution{Inria Paris-Rocquencourt}
  \city{Rocquencourt}
  \country{France}
}
\author{John Smith}
\affiliation{\institution{The Th{\o}rv{\"a}ld Group}}
\email{jsmith@affiliation.org}

\author{Julius P. Kumquat}
\affiliation{\institution{The Kumquat Consortium}}
\email{jpkumquat@consortium.net}

\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
With increasing complexity of hardwares, systems with different memory nodes are ubiquitous in high performance computing (HPC). It is paramount to develop strategies to overlap the data transfers between memory nodes with the computation in order to exploit the full potential of these systems. In this article, we consider the problem of deciding the  order of data transfers between two memory nodes for a set of independent tasks  with the objective to minimize the makespan. We propose an optimal heuristic to obtain the order of data transfers when there is not any restriction on memory limit. We also show that with finite memory capacity obtaining the optimal order of data transfers is an NP-complete problem. We propose several heuristics for this problem and provide details about their favorable situations. We present analysis of our heuristics on traces, obtained by running 2 molecular chemistry kernels, namely, Hartree–Fock (HF)  and Coupled Cluster Single Double (CCSD) on 10 nodes of an HPC system. Our results show that some of our heuristics achieve significant overlap for moderate memory capacities and are very close to the lower bound of makespan.
\end{abstract}

%%\begin{CCSXML}
%%	<ccs2012>
%%	<concept>
%%	<concept_id>10010520.10010553.10010562</concept_id>
%%	<concept_desc>Computer systems organization~Embedded systems</concept_desc>
%%	<concept_significance>500</concept_significance>
%%	</concept>
%%	<concept>
%%	<concept_id>10010520.10010575.10010755</concept_id>
%%	<concept_desc>Computer systems organization~Redundancy</concept_desc>
%%	<concept_significance>300</concept_significance>
%%	</concept>
%%	<concept>
%%	<concept_id>10010520.10010553.10010554</concept_id>
%%	<concept_desc>Computer systems organization~Robotics</concept_desc>
%%	<concept_significance>100</concept_significance>
%%	</concept>
%%	<concept>
%%	<concept_id>10003033.10003083.10003095</concept_id>
%%	<concept_desc>Networks~Network reliability</concept_desc>
%%	<concept_significance>100</concept_significance>
%%	</concept>
%%	</ccs2012>
%%\end{CCSXML}
\ccsdesc[100]{Computer systems organization~Embedded systems1}
\ccsdesc[100]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

\keywords{Communication Scheduling, Memory Nodes, Runtime Systems, Communication-Computation Overlap, Molecular Chemistry}

\maketitle

@@

	\begin{figure}[htb]
		\centering
		\includegraphics[scale=0.35]{Figs/original_swapped_schedules}
		\caption{ \label{fig:bothSchedule} Original and Swapped Schedules.}
	\end{figure}
	\todo[inline]{SKumar: decrease space between both schedules of Figure~\ref{fig:bothSchedule}}
	We write the following constraints based on the two schedules of the Fig~\ref{fig:bothSchedule}.

@@

%%	\begin{table*}[htb]
		\begin{tabular}{ |c|c|c| }
			\hline
			Task & Input Transfer time ($comm_i$)& Computation time ($e_i$)\\ \hline
			$K_0$ & $0$ & $3$ \\ \hline
			$K_1, \cdots, K_{m-1}$ & $b'=b+3x$ & $3$\\ \hline
			$K_m$ & $b'=b+3x$ & $0$ \\ \hline
			$1\le i \le 3m, A_i$ & $1$ & $a_i' = a_i +x$\\ \hline
		\end{tabular}
%%	\end{table*}
	Where $x= 2 max\{a_i:1\le i\le 3m\}$\\ 
	Memory capacity $C=b'+3$ and $L=m(b'+3)$, this completes the construction of instance $I$ of the 
	problem $DT$.

@@

	Conversely, if such a partition of $A$ is achievable, then we can construct a feasible schedule $S$ 
	without idle times by the pattern depicted in Fig~\ref{fig:firstSegment}. We have shown that $3PAR$ 
	problem reduces to $DT$, and hence problem $DT$ is NP-complete.
	\todo[inline]{Skumar: Increase width of both channels of Figure~\ref{fig:firstSegment}.}

	\begin{figure}[htb]
		\centering
		\includegraphics[scale=0.5]{Figs/first_segment}
		\caption{ \label{fig:firstSegment} First segment of Schedule $S$.}
	\end{figure}

@@

\begin{table*}[htb]
	\scriptsize
	\begin{tabular}{|c|p{3cm}|}
		\hline
		\textbf{Heuristic} & \textbf{Favorable Situation} \\ \hline
		order\_of\_optimal\_strategy\_infinite\_memory & Unlimited memory capacity (\textcolor{green}{Optimal}) \\ \hline
		increasing\_order\_of\_communication\_strategy & Unlimited memory capacity and tasks are compute intensive (\textcolor{green}{Optimal}) \\ \hline
		decreasing\_order\_of\_computation\_strategy & Unlimited memory capacity and tasks are communication intensive (\textcolor{green}{Optimal}) \\ \hline
		increasing\_order\_of\_communication\_plus\_computation\_strategy & Moderate memory capacity and most tasks are compute intensive \\ \hline
		decreasing\_order\_of\_communication\_plus\_computation\_strategy & Moderate memory capacity and most tasks are communication intensive \\ \hline
		largest\_communication\_task\_respects\_memory\_restriction &  Limited memory capacity and significant percentage of tasks are communication intensive \\ \hline
		smallest\_communication\_task\_respects\_memory\_restriction & Limited memory capacity and most tasks are compute intensive \\ \hline
		maximum\_accelerated\_task\_respects\_memory\_restriction &  Limited memory capacity and number of communication and compute intensive tasks are approximately equal \\ \hline
		optimal\_order\_infinite\_memory\_largest\_communication\_task\_respects\_memory\_restriction & Moderate memory capacity and significant percentage of tasks are communication intensive\\ \hline
		optimal\_order\_infinite\_memory\_smallest\_communication\_task\_respects\_memory\_restriction & Moderate memory capacity and most tasks are compute intensive \\ \hline
		optimal\_order\_infinite\_memory\_maximum\_accelerated\_task\_respects\_memory\_restriction & Moderate   memory capacity and number of communication and compute intensive tasks are approximately equal \\ \hline
	\end{tabular}\caption{~\label{tab:heuristicsAndFavorableScenarios}Heuristics and their favorable scenarios}
\end{table*}

@@

Figures ~\ref{fig:ratio_to_optimal_hf} and ~\ref{fig:ratio_to_optimal_ccsd} depict the distribution of the performance  of each heuristic for all considered memory capacities, where plots are categorized by memory capacities. For each memory capacity and each heuristic, the box on the plot displays the median, first and last quartile, and the whiskers indicate minimum and maximum values, with outliers are shown by black dots.
\subsection{HF Performance}	
\begin{figure*}[htb]
	\includegraphics[scale=0.25]{./results/plots/ratio_to_optimal_selected_hf.pdf}
	\caption{Comparison of different heuristics for HF.}
	\label{fig:ratio_to_optimal_hf}
\end{figure*}
As indicated above, HF tasks are highly homogeneous, this is also noticeable in Figure~\ref{fig:ratio_to_optimal_hf}. All heuristics depict similar behavior for minimum memory capacity $m_c$  and  increasing the memory capacity slightly does not change the performance of all heuristics. As memory capacity increased further, dynamic variants of heuristics starts performing better and for the moderate memory capacities (close to $2m_c$ ), static order with dynamic correction variants outperform others. \textit{Gilmore-Gomory} heuristic does not achieve good performance. it can be understand from the fact that the task sequence is obtained considering no extra memory is available, but here we apply the same sequence for a different scenario,  overall memory consumed at each point is limited to memory capacity.

\begin{figure}[htb]
	\includegraphics[scale=0.15]{./results/plots/inverse_ratio_to_optimal_hf-best.pdf}
	\caption{Comparision of best variants of all categories for HF.}
	\label{fig:ratio_to_optimal_best_hf}
\end{figure}

@@

\begin{figure*}[htb]
	\includegraphics[scale=0.25]{./results/plots/ratio_to_optimal_selected_ccsd.pdf}
	\caption{Comparison of different heuristics for CCSD.}
	\label{fig:ratio_to_optimal_ccsd}
\end{figure*}

@@

\begin{figure}[htb]
	\includegraphics[scale=0.15]{./results/plots/inverse_ratio_to_optimal_ccsd-best.pdf}
	\caption{Comparision of best variants of all categories for CCSD.}
	\label{fig:ratio_to_optimal_best_ccsd}
\end{figure}

@@

\subsection{MILP Based Strategy}
We use GLPK solver v4.65 to solve the MILP formulation described in Section~\ref{subsec:linearprogrammingformulation}. The solver was unable to handle number of variables required to solve MILP at the scale of our interest. Hence, we solve the linear program iteratively for a small subset of size $k=3,4,5$ (solver was unable to handle number of variables required for $k\ge6$ ) and represent the makespan calculated by this heuristic as $lp.k$. At the boundary of two iterations we fixed the event (communication or computation) of an unfinished task who has  started before the boundary point and consider other events flexible. We compute various  $lp.k$ values for different memory capacities and observe that most of the other heuristics perform better than this heuristic. Figure~\ref{fig:iterativeLpSolution} shows the comparison of different heuristics with MILP based heuristic for various memory capacities of a single trace file.
\todo[inline]{SKumar: Obtain MILP based makespans for new memory capacities}
\begin{figure}[htb]
	\includegraphics[scale=0.5]{./results/makespan_with_lp.pdf}
	\caption{Comparision of different heuristics with MILP solution based heuristic.}
	\label{fig:iterativeLpSolution}
\end{figure}

@@

\begin{figure*}[htb]
	\subfloat[Best variants of HF.\label{fig:hf_best_variants_batch}]{%
		\includegraphics[width=.45\linewidth]{./results/plots/inverse_ratio_to_optimal_hf_batch-best.pdf}
	}%\\
	\subfloat[Best variants of CCSD.\label{fig:ccsd_best_variants_batch}]{%
		\includegraphics[width=.45\linewidth]{./results/plots/inverse_ratio_to_optimal_ccsd_batch-best.pdf}
	}%\\
	\caption{Best variants of all categories where heuristics are applied in the batches of 100 tasks.}
	\label{fig:best_variants_batch}
\end{figure*}

@@

\subsection{Impact of heterogeneity on different heuristics}
\todo[inline]{SKumar: remove this subsection or replace with an experiment which shows the performance difference with all classes of heuristics for randomly generated inputs.}
\begin{figure*}[htb]
	\includegraphics[scale=0.25]{./results/plots/ratio_to_optimal_random_selected_ccsd.pdf}
	\caption{Comparison of different heuristics for CCSD with highly heterogeneous tasks.}
	\label{fig:ratio_to_optimal_ccsd_random}
\end{figure*}

@@

\begin{acks}
Identification of funding sources and other support, and thanks to individuals and groups that assisted in the research and the preparation of the work should be included in an acknowledgment section, which is placed just before the reference section in your document.

\end{acks}

\section{Acknowledgments}

%
% The next two lines define the bibliography style to be used, and the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{communications}
